---
title: "Data Storage Formats"
author:
  - name: Louis M Penrod, Ph.D.
date-modified: last-modified
---

This document will go over the many, many ways data can be stored and how to read them into R. This will be a living document; growing as time goes on to explore topics more in depth, add new methods of reading in data, and add new data formats that can be read in.

# Text: Free-form

Free-form text files are those that don't have a systematic structure. They may have a systematic structure if they are being generated by the same process of software, but there is rarely similarities between processes.

Often, these are meant for visual representation of data, like outputs of statistical results or listings, rather than data storage, but often these contain important information that needs to be parsed out.

There is not too much advice I can give here that applies to every free-form text file because one is going to be different. 

There may be some standardization if a particular software is writing the files (e.g., modeling software writing standard outputs) in which case it would be best to write a custom function, or find one online if it exists, to parse that file.

Start with one of these functions below:

```{r}
#| eval: false

readLines()

#or

readr::read_lines()
```

I usually just use `readLines()` because it reduces the need for a dependency (the `{readr}` package), but `read_lines()` does have some additional functionality like the ability to skip a certain number of rows.

One example may be trying to parse information from a text file that is the output of some statistical software. Here, I've just run a quick ANOVA with some assumptions tests and written the output to a text file.

```{r}
lines <- readLines("./data_storage_formats/mtcars_test_results.txt")

cat(lines,sep = "\n")
```

Now let's say I just want the F statistic and p-value from the ANOVA. I need a way to extract just that info. 

Here is how I would tackle this, assuming a somewhat standard output from whatever created this file.

I'd start by identifying the line number that contains the section header for the ANOVA. From that, I know the F statistic and p-value will be two lines below that

```{r}
line_num_aov <- which(grepl(pattern = "ANOVA", lines)) 
line_num_targ <- line_num_aov + 2

line_num_targ
```

Now I need a way to extract just the components from that line. 
There is nothing within that line to tell me where the values I want are. For example, the Bartlett's K-squared value follows a label that says that and proceeds a comma. I could use that as clear rules for extracting that value. We don't have that for the ANOVAs F statistic and p-value.

Instead, I think I'm going to have to base it off parts. If we split that line into parts based off spaces between values, and the structure is always the same, I know the F statistic and p-value are the 5th and 6th values [1 = variable label, 2 = Df, 3 = Sum Sq, 4 = Mean Sq, 5 = F value, 6 = Pr(>F), 7 = Signif code].

So I'll split on spaces and extract the parts I need.

```{r}
line_split <- strsplit(lines[line_num_targ], split = "\\s+")[[1]]
f_val <- as.numeric(line_split[5])
p_val <- as.numeric(line_split[6])
```

```{r}
f_val
p_val
```

There you have it, the F statistic and p value from the ANOVA.


There are some special cases of free-form text files which can include loggers and listings depending on how they have been set up.

## Loggers

Some loggers output free-form text, but others could be fixed-width or even delimited. For the free-form versions, again, you are going to need a custom parser. I do find these easier than listings because they usually don't wrap records onto multiple lines as there is no width restrictions like listing have.

I have some water quality logger data for a project I worked on which I've taken any identifying information out of (although the data is out in the public). This will be a paired down version without some checks and processing steps which may not be needed in your logger data.

```{r}
library(stringr)
lines <- readLines("./data_storage_formats/ex_free_form_text.txt")
```

```{r}
#| results: hold

cat(head(lines, n = 20), sep = "\n")
cat("...\n")
cat(tail(lines, n = 20), sep = "\n")
```

First step I took was to identify the different parts in the top section using the line numbers.

```{r}
date_pattern <- "\\b\\d{2}/\\d{2}/\\d{4}\\b"
   
# find the first row with the date as this is the first row with data 
header <- which(grepl(date_pattern, lines))[1] - 2
# get the variables
headers <- toupper(stringr::str_split(lines[header], "\\s{2,}")[[1]])

# get the dashed line which separated the data and the variable names  
dash_line <- lines[header + 1]

# Find the start and end of each series of dashes
# These defined the edges of the columns
matches <- stringr::str_match_all(dash_line, "(-+)")
dash_spans <- matches[[1]]
# Use gregexpr to get start positions of each dash group
starts <- unlist(gregexpr("-+", dash_line))
breaks <- c(starts[-1]-1, nchar(dash_line))
```

Next, I grabbed the quality codes at the end of the file

```{r}
# get first like of quality section
quality_start <- which(stringr::str_detect(lines, "Quality Codes Key:")) - 1

# get all rows that had a number, dash, then something following
quality_lines <- which(stringr::str_detect(lines, "\\d+\\s-\\s.*"))
quality_lines <- quality_lines[quality_lines > quality_start]
quality_lines <- lines[quality_lines]

# split into two columns based on the dash
# and make into a data frame
quality_lines <- stringr::str_split(quality_lines, 
                                    pattern = "\\s-\\s", 
                                    n = 2, 
                                    simplify = TRUE)
quality_lines <- as.data.frame(quality_lines)
colnames(quality_lines) <- c("QUALITY", "QUALITY_KEY")
```


Next, reduce down to just the lines for the data and split based on the positions we identified in the previous step.

```{r}
lines <- lines[grepl(date_pattern, lines)]
lines <- lines[!grepl("---{4,}", lines)]

df <- purrr::map(lines, function(line){
        
        x <- purrr::map2(starts, breaks, ~ str_trim(str_sub(line, .x, .y))) %>% 
          as.data.frame()
        colnames(x) <- headers
        return(x)
        
      }) %>% 
        dplyr::bind_rows()

rownames(df) <- NULL

```

Add the quality code to the dataset and clean it up

```{r}
df <- dplyr::left_join(df, quality_lines, by = "QUALITY") %>% 
   dplyr::mutate(dplyr::across(dplyr::where(is.character), ~stringr::str_replace(.x, "\"\"", ""))) %>% 
  dplyr::mutate(QUALITY = as.numeric(QUALITY))
```

This gets us the final dataset:

```{r}
#| echo: false
#| results: hold
head(df)
cat("\n...\n\n")
tail(df)
```


## Listings

Listings are a whole different discussion. I really do not understand how, in the 21st century, the best way to communicate record or even subject level data is to print those values in a report or document. I don't believe anyone is actually looking through each value. Providing the actual data in a format that could be better explored (yes, even Excel which has filters) would be better. If it was just printing the data, I would not have such a problem with it, but on occasion, I've been told a listing was the only format someone had the data...

Several reasons I dislike these as data:

- Data is too large to print all in one document, both too many columns and rows. The data has to be split into panels. These will have to be parsed out and recombined in both directions.  
- Sometimes, longer text within a "cell" is wrapped so it is technically in multiple rows.  
- Columns can vary in width so parsing based on fixed-width is not possible.
- Generally there are footnotes or counts every page that has to be ignored, but breaks up the tabular structure.  
- Values in columns are often grouped, similar to merged cells in Excel rather on every row they apply to.  
- Since it is usually visual, decode or placeholders are included rather than the actual values in the data. Not usually a huge issue but can cause issues if it mixed class types.  
- Formatting is never consistent. Every tool that produces a listing, produces a different structure. This means there is no one way that can be used to parse out the data.


Honestly, R is not my first try to parse these into something usable. If it is a PDF, the export to Excel works half-decent and you can then manually deal with all of the other issues manually. For Word, copy into Excel and, again, deal with it there. Since ever listing you get will have a different structure, I'm not sure it is work the effort to code up a parsing solution.



# Text: Self-describing

## JSON

## YAML

## XML

## HTML

# Text: Delimited

## CSV

## TSV

## Other separated text file

# Text: Layout-dependent

## Fixed-width

# Spreadsheets

## Excel (.xlsx, .xls)

## .ods

## Google Sheets

# R-native

## RDS

## RData

## QS/QMS

# Statistical software

## SAS7BDAT/SAS7BCAT

## XPT

## SPSS

## Stata

# Documents

## Word

## RTF

## PDF (text-based or scanned)

# Images and Videos

## Photos

## Video frames

## EXIF metadata

# Databases

## DuckDB

## Other DBs

## SQLite

## MS Access

# Cloud warehouse/lakes

## Databricks

## Snowflake

# Spatial

## Raster

## NetCDF

## Shapefiles

## GeoPackage

## GeoJSON

## KML

# Binary/columnar

## Parquet
## Feather/Arrow
## HDF5

# APIs

## Raw API pulls
## R packages that hit APIs and format

# Other languages' formats

## MATLAB
## Python pickle/NumPy

# Domain specific

There are so many formats that are only used in specific domains and are less common. 

These could be things like signal/audio, DICOM, or types specifically for things like genomics.

